### 2022-08-05 chapter6/chapter.tex
ch 7/8 comments

Lastly, for the DG scheme I know of no rigorous argument which shows that we can use the DG approximation to the generator of a fluid queue to approximate performance measures of fluid-fluid queues, such as the operator \(\mathbb \Psi\). Here, we rely on the intuition that substituting the DG approximation for the true operator \emph{should} give reasonable results, and further, \emph{assume} that the resulting approximations are polynomial. For example this means that, other than being the solution to a matrix Riccati equation, the DG approximations to \(\mathbb \Psi\) that we compute in this chapter are not actually rigorously defined.

On the other hand, using the stochastic approximations (QBD-RAP of uniformisation), results in rigorously defined operators as proved in \cite{bgnp2021}.

In summary, the QBD-RAP and uniformisation schemes are the only two schemes which result in rigorously defined approximations, preserve the axioms of probability, and, of the two, the numerical experiments demonstrate that the QBD-RAP methods is superior.

@@

The main focus of the numerical experiments is on the error as the number of basis functions per cell is increase. In the numerical experiments the number of basis functions used to approximate the solutions is the same on all cells and the number of basis functions on a cell (and hence all cells) is referred to as the \emph{dimension} of the scheme. For the discontinuous Galerkin scheme, increasing the dimension of the scheme means increasing the number of polynomial basis functions used to approximate the solution within each cell. For example, if we use 3 basis functions in the discontinuous Galerkin scheme, we approximate the solution by 3 linearly independent quadratics on each cell and hence the approximation is a quadratic on each cell. For the QBD-RAP scheme, increasing the dimension of the scheme means increasing the order of the CME distribution used to construct the scheme. To make a comparable equivalent for the uniformisation scheme, we divide each cell into smaller sub-cells over which we approximate the solution. That is, for a dimension \(p\) uniformisation scheme we divide each cell into \(p\) sub-cells. Equivalently, we may think of a dimension \(p\) uniformisation scheme as using \(p\) piecewise constant functions to approximate the solution on each cell. For all schemes (DG, QBD-RAP and uniformisation), if we construct a dimension \(p\) approximation, there are \(K\) cells, \(N\) phases and \(c_i\neq 0\) for all \(i\in\calS\), then the resulting approximation to the generator \(\mathbb B\) is a square matrix of dimension \(pKN + N\).\footnote{There are \(pK\) basis functions for each phase which approximate the solution on the interior of the domain, hence \(pKN\) basis functions in total, plus an additional \(N\) dimensions to capture boundary conditions.} In this sense each approximation scheme leads to matrices of the same order (although not necessarily the same number of non-zero elements).

We implement the discontinuous Galerkin scheme with and without a slope-limiter (\cite{c99}, \cite{nodalDGBook}~Section~5.6.2, see also Section~\ref{sec: slope limiting}). The slope limiter stops oscillations appearing in the approximations and hence also avoids non-probabilistic approximations to PDFs and CDFs. We investigate two slope-limited schemes. In the scheme we refer to the \emph{DG-lim} scheme, we use a DG scheme of dimension \(p\) and apply the Generalised MUSCL slope limiter as described in Section~\ref{sec: slope limiting}. In the \emph{DG-lin-lim} scheme, we follow the construction described in Section~\ref{sec: anoth pos pres}, which is effectively a DG scheme with cells of width \(2\Delta/(p+1)\), for \(p=1,3,5,...,21\), and a basis of dimension \(2\) (linear) on each cell, with an upwind flux, and we apply the Generalised MUSCL limiter. Thus, the DG-lin-lim scheme uses \(p+1\), \(p=1,3,5,...,21\), basis functions to represent the solution on each of the original cells \(\calD_k\), and we define the dimension of the scheme to be \(p+1\). In the presence of discontinuities, slope-limiting forces the approximation to be linear, i.e.~an order~1 polynomial, around the discontinuity. For DG schemes of dimension greater than 2 with a slope limiter and in the presence of discontinuities, the ability of the scheme to represent higher-order polynomials is not well-utilised as the slope limiter forces the solution to be linear anyway. The DG-lin-lim scheme recognises this and instead of using higher-order polynomials, uses piecewise-linear functions and smaller cell widths to approximate the solution.

The concept of slope-limiting originates in the context of numerically integrating (typically with respect to a \emph{time} variable) a so-called \emph{semi-discrete system}, as we do in Section~\ref{sec: transient approx}. However, approximating the operator \(\mathbb\Psi\) of a fluid-fluid queue amounts to solving an implicit matrix-equation and there is no concept of numerical integration with respect to time. Hence, the concept of slope-limiting, it is not directly transferable. One could process the initial condition and post-process the solution with a limiter. However, there are possibly better choices for this post-processing, such as filtering, \citep[Section~5.6.1]{nodalDGBook}. One could also view the composition of the DG operator and the limiting operator as a one, but the result is a non-linear operator which is of no use for simplifying the Riccati Equation which characterises \(\mathbb \Psi\). Considering there is no theory on the use of the DG scheme to approximate \(\mathbb \Psi\) and hence the resultant approximation is not rigorously-defined, it is not known if the use of a limiter in this context is possible.

@@

For schemes which require us to integrate over time we do not investigate the stability of the schemes with respect to the \(t\)-step-size, or the time-integration scheme itself (where required). Instead, we fix the time-integration step size for each numerical experiment at a suitably small value to obey a certain stability criterion (a CFL-like condition, \citep[Section~4.8]{nodalDGBook}). Moreover, we always implement the strong stability preserving Runge-Kutta method of order 4 with 5 stages (\cite{sr2002}, \cite{nodalDGBook}~Section~5.7, see also Section~\ref{sec: time integration}), which claims to introduce no more oscillations into the solution as we integrate over time.

@@

The structure of this chapter is as follows. In Section~\ref{sec: recon num} we compute approximations to various initial conditions for the different schemes and observe their performance which allows us to investigate the performance of the reconstruction methods without considering a specific model or any dynamics. In Section~\ref{sec: wave num} we investigate travelling wave problems with various initial conditions. For the travelling wave problem the dynamics are deterministic, which means that the solution is known, and we do not need to resort to simulation as a \emph{ground-truth}. Furthermore, the travelling wave problems allow us to investigate the ability of the schemes to approximate the flow of probability across cells without any stochastic dynamics. Next, we investigate a simple fluid queue with two phases. Section~\ref{sec:stat} investigates approximations to the limiting distribution of the queue, Section~\ref{sec: transient approx} investigates approximations to the transient distribution of the queue for two initial conditions, and Section~\ref{sec: hit approx} investigates approximations the first hitting time of the fluid level on the set \(\{0,1\}\). Lastly, we apply the approximation methods to a fluid-fluid queue in Section~\ref{sec: ffq num}, compute approximations to \(\mathbb \Psi\) and ultimately, approximate the distribution of \(\{(X(t),\varphi(t))\}\) at the time at which \(\{Y(t)\}\) first returns to \(0\).

@@

\subsection{QBD-RAP closing operators}\label{sec: closing vecs numerics}
The three closing operators we investigate are the following. The unnormalised closing vector, 
\begin{align}
	\bs a {\bs v}(x) &= \bs a e^{\bs Sx}\bs s.\tag{\ref{eqn: alal}}
	%\\\bs a {\bs v}_-(x-y_\ell) &= \bs a e^{\bs S(x-y_{\ell})}\bs s,\label{eqn:ndaenf}
\end{align} 
A normalised version of the closing operator in Equation~(\ref{eqn:546}), with the normalised version given by 
\begin{align}
	\bs a {\bs v}(x) &= \cfrac{\bs a \left(e^{ \bs{S}x} \bs s + e^{ \bs{S}(2\Delta-x)} \bs s\right)}{1-\bs a e^{\bs S 2\Delta}\bs e}. \tag{\ref{eqn: nonlin}}
\end{align} 
We refer to the closing operator in (\ref{eqn: nonlin}) as the \emph{non-linear normalised} closing operator.\footnote{Technically we have not yet proved that the \emph{non-linear} normalised closing operator leads to a convergent approximation scheme.} The third closing vector we investigate is given by 
\begin{align}
	\bs a {\bs v}(x) &= \bs a \left(e^{ \bs{S}x} \bs s + e^{ \bs{S}(2\Delta-x)}\right)\left[I-e^{S2\Delta}\right]^{-1} \bs s,\tag{\ref{eqn:5462}}
\end{align}
which we refer to as the \emph{normalised} closing operator.

@@

	Figure~\ref{fig: fun 2 ks error qbdrap closing vecs} also shows that the non-linear normalised and normalised reconstructions perform very similarly -- this is the case throughout much of this section. The difference between the non-linear normalised and normalised reconstructions is how they treat mass in the tail of a matrix exponential distribution (from \(2\Delta\) onward). Intuitively, there should be very little mass in the tail of the distribution as we are using concentrated matrix exponential distributions in the reconstruction.

@@

Some insight is gained by looking at Figure~\ref{fig: pdf reconstructed} where we plot the reconstructed PDFs for the unnormalised (\ref{eqn:546}) and normalised (\ref{eqn:5462}) closing operators for dimension 1, 3, 5 and 7, as well as the true PDF. Observing Figure~\ref{fig: pdf reconstructed} notice that the unnormalised reconstruction fails to capture the density at the left of each of the plots. This feature is due to a significant amount of mass being lost due to the truncation. In comparison, the reconstruction with the normalised closing operator is much better in this region due to the `folding' around \(\Delta\) in the construction of the closing operator. The `folding' in the normalised closing operator manifests itself as more mass at the left of these plots compared to the unnormalised reconstructions. % Recall that, as discussed in Section~\ref{}, for the reconstruction we must to allocate the initial condition to a phase with positive rate, or negative rate, so that we can use the appropriate approximation and reconstruction method as discussed in Section~\ref{sec: closing}. Here we suppose that the phase is positive, so the `folding' around \(\Delta\) in the normalised closing operators appears appears at the left of the plots. In general, reconstructions via the unnormalised closing operator typically underestimate the value of the function in this region.

\begin{figure}[h]
	\centering
	\includegraphics[width=\textwidth,trim={0cm 1.25cm 0cm 1.25cm},clip]{chapter6/figs/qbdrap_closing_vec/fun4/pdfs_formatted.pdf}
	\caption{Approximations to the PDF \(f(x)=1,\, x\in[0,1)\) (orange dotted line) for Example~\ref{ex: al}, constructed using matrix exponential distributions of various dimensions, and using the unnormalised closing operator (blue solid line) and normalised closing operator (green dash-dotted line).}
	\label{fig: pdf reconstructed}
\end{figure} 
%\exampleFloatBarrier
Figure~\ref{fig: pdf reconstructed} also show that both closing operators fail to approximate the initial condition well at the right-hand side of the interval. A potential area for further research is to determine if there is a different reconstruction method or another closing operator which could alleviate this issue. 
\end{example}

\begin{example}\label{ex: aasmm}
	\begin{figure}[h]
		\centering
		\includegraphics[width=0.5\textwidth,trim={0.5cm 0.8cm 0.2cm 1.25cm},clip]{chapter6/figs/qbdrap_closing_vec/fun6/ks_error_formatted.pdf}%
		\includegraphics[width=0.5\textwidth,trim={0.5cm 0.8cm 0.2cm 1.25cm},clip]{chapter6/figs/qbdrap_closing_vec/fun6/l2_pdf_error_formatted.pdf}
		\caption{KS error (left) and \(L^2\) error for the PDFs (right) of Example~\ref{ex: aasmm}, for the three closing vectors considered; unnormalised (blue solid line), non-linear normalised (orange dashed line) and normalised (green dash-dotted line). The error curves for the non-linear normalised (orange) and normalised closing vectors are almost coincident. The black dotted lines are linear least-squares fits to the last 8 data points and the slopes of the least square lines are written next to the last point.}
		\label{fig: fun 6 ks error qbdrap closing vecs}
	\end{figure}
	Next we consider approximating the initial distribution with density \(-6x^2+6x\) on \([0,1)\). Observing the left-hand panel of Figure~\ref{fig: fun 6 ks error qbdrap closing vecs} which plots the KS error against the dimension of the reconstruction for the three closing operators, we once again see that the reconstruction using the unnormalised closing operator performs the worst, while the performance of the two normalised reconstructions is indistinguishable. However, if we instead look at the right-hand panel of Figure~\ref{fig: fun 6 ks error qbdrap closing vecs} which plots the \(L^2\) error between the reconstructed PDF and the true PDF, then the unnormalised closing operator performs better than the two normalised ones for dimension 5 and above.

	\begin{figure}[h]
		\centering
		\includegraphics[width=\textwidth,trim={0cm 1.25cm 0cm 1.25cm},clip]{chapter6/figs/qbdrap_closing_vec/fun6/pdfs_formatted.pdf}
		\caption{Approximations to the PDF \(f(x)=-6x^2+6x\) (orange dotted line) for Example~\ref{ex: aasmm}, constructed using matrix exponential distributions of various dimensions, and using the unnormalised closing operator (blue solid line) and normalised closing operator (green dash-dotted line).} 
		\label{fig: pdf reconstructed quadratic}
	\end{figure} 
	%\exampleFloatBarrier
The fact that the unnormalised closing operator outperforms the two normalised ones can be explained by the `folding' of the normalised operators around \(\Delta\). In Figure~\ref{fig: pdf reconstructed quadratic} we plot the unnormalised and normalised reconstructions along with the true PDF, \(-6x^2+6x\). In Figure~\ref{fig: pdf reconstructed quadratic} at the left of the plots, we observe that the normalised reconstructions over-estimate the density function, whereas the unnormalised reconstruction looks to be doing better. The `folding' in the normalised closing operator manifests itself as more mass at the left of these plots compared to the unnormalised reconstructions. 
\end{example}

\begin{example}\label{ex: kjfk}
	\begin{figure}[h]
		\centering
		\includegraphics[width=0.5\textwidth,trim={0.5cm 0.8cm 0.2cm 1.25cm},clip]{chapter6/figs/qbdrap_closing_vec/fun7/ks_error_formatted.pdf}%
		\includegraphics[width=0.5\textwidth,trim={0.5cm 0.8cm 0.2cm 1.25cm},clip]{chapter6/figs/qbdrap_closing_vec/fun7/l2_pdf_error_formatted.pdf}
		\caption{KS error (left) and \(L^2\) error of the PDF (right) for Example~\ref{ex: kjfk} for the three closing vectors considered; unnormalised (blue solid line), non-linear normalised (orange dashed line) and normalised (green dash-dotted line). The error curves for the non-linear normalised (orange) and normalised closing vectors are almost coincident. The black dotted lines are linear least-squares fits to the last 8 data points and the slopes of the least square lines are written next to the last point.}
		\label{fig: fun 7 ks error qbdrap closing vecs}
	\end{figure}
	%\exampleFloatBarrier
	Lastly, we consider approximating the initial distribution with PDF \(3e^{-3x}/(1-e^{-3})\). This density function is at a maximum at the left of the region. Considering what we have learnt so far about the unnormalised operator underestimating in this region, we expect that the unnormalised closing operator will perform relatively poorly in this case. Indeed, observing Figures~\ref{fig: fun 7 ks error qbdrap closing vecs} we see that this is the case; the normalised reconstructions perform relatively well compared to the unnormalised reconstruction, as measured by both error metrics (KS statistic and \(L^2\) norm on the PDF). Plots of the PDFs (omitted), confirm that this is due to the loss of mass at the left-hand side of the region due to the truncation. 
\end{example}

Table~\ref{table: error rates for QBD-RAP closing} summarises the rates of convergence of the \(L^2\) error on the PDFs for unnormalised and normalised closing operators in the examples above (i.e.~the slopes of the black dotted lines in the error plots). In general, the normalised closing operators converge at a rate approximately as fast, or faster, than the unnormalised reconstruction.
\begin{table}
	\centering
	\begin{tabular}{c|c|c|c}
			Example & PDF on \([0,1)\) & \(\beta_1\), unnormalised & \(\beta_1\), normalised \\\hline
			\ref{ex: another example} & \(2\times 1(x<0.5)\) & -1.07 & -0.98 \\
			\ref{ex: al} & 1 & -0.99 & -0.91 \\
			\ref{ex: aasmm} & \(-6x^2+6x\) & -2.72 & -2.57 \\ 
			\ref{ex: kjfk} & \(3e^{-3x}/(1-e^{-3})\) & -1.11 & -2.35 
	\end{tabular}
	\caption{\label{table: error rates for QBD-RAP closing} Estimated rates of convergence, \(\beta_1\), of the \(L^2\) error on the PDFs for unnormalised and normalised closing operators in the examples above (i.e.~the slopes of the black dotted lines in the error plots).}
\end{table}

@@

\begin{example}\label{ex: exa}
	\begin{figure}[h]
		\centering
		\includegraphics[width=0.5\textwidth,trim={0.5cm 0.8cm 0.2cm 1.25cm},clip]{chapter6/figs/comp/fun1/meshs_ks_error_formatted.pdf}%
		\includegraphics[width=0.5\textwidth,trim={0.5cm 0.8cm 0.2cm 1.25cm},clip]{chapter6/figs/comp/fun1/meshs_l1_cdf_error_formatted.pdf}
		\caption{KS error (left) and \(L^1\) error of the CDF (right) of Example~\ref{ex: exa} for the DG (blue solid line), uniformisation (orange dashed line) and QBD-RAP (purple dashed line) schemes. The black dotted lines are linear least-squares fits to the last 8 data points and the slopes of the least square lines are written next to the last point.}
		\label{fig: fun 1 comp} 
	\end{figure}
	First we consider the initial condition with CDF \(1(x\geq 0.5)\), that is, a point mass of \(1\) at \(0.5\). This distribution does not have a PDF, so we compare the CDFs only. In Figure~\ref{fig: fun 1 comp} we plot the KS metric (left) and \(L^1\) metric (right) between the true CDF and the reconstructed approximations. Observing the KS metric, it appears that none of the schemes converge and the KS error sits around 0.5 for all schemes. This reflects the fact that convergence in distribution implies point-wise convergence of the CDFs except, perhaps, at points of discontinuity. None of the schemes converge at the discontinuity at \(x=0.5\). Observing the \(L^1\) error between the true CDF and reconstructed approximation (which is the area between the two CDFs) we now see that the schemes appear to show the convergent behaviour we expect. Here, the uniformisation scheme appears to perform the best, while the QBD-RAP scheme performs the worst. The rate of convergence of the QBD-RAP scheme appears to be similar to the rate of convergence of the DG scheme.

>>>>

Although we have been unable to supply a rigorous argument to show that we can use the DG approximation to the generator of a fluid queue to approximate performance measures of fluid-fluid queues, such as the operator \(\mathbb \Psi\), we can rely on the intuition that substituting the DG approximation for the true operator \emph{should} give reasonable results, and further, \emph{assume} that the resulting approximations are polynomial.

On the other hand, using the stochastic approximations (QBD-RAP or uniformisation), results in rigorously defined operators as proved in \cite{bgnp2021}.

In summary, the QBD-RAP and uniformisation schemes are the only two schemes which result in rigorously defined approximations, preserve the axioms of probability, and, of the two, the numerical experiments demonstrate that the QBD-RAP method is superior.

@@

The main focus of the numerical experiments is on the error as the number of basis functions per cell is increased. In the numerical experiments the number of basis functions used to approximate the solutions is the same on all cells and the number of basis functions on a cell (and hence all cells) is referred to as the \emph{dimension} of the scheme. For the discontinuous Galerkin scheme, the dimension of the scheme means is the number of polynomial basis functions used to approximate the solution within each cell. For example, if we use 3 basis functions in the discontinuous Galerkin scheme, we approximate the solution by 3 linearly independent quadratics on each cell and hence the approximation is a quadratic on each cell. For the QBD-RAP scheme, the dimension of the scheme is the order of the CME distribution used to construct the scheme. To make a comparable equivalent for the uniformisation scheme, we divide each cell into smaller sub-cells over which we approximate the solution. That is, for a dimension \(p\) uniformisation scheme we divide each cell into \(p\) sub-cells. Equivalently, we may think of a dimension \(p\) uniformisation scheme as using \(p\) piecewise constant functions to approximate the solution on each cell. For all schemes (DG, QBD-RAP and uniformisation), if we construct a dimension \(p\) approximation, there are \(K\) cells, \(N\) phases and \(c_i\neq 0\) for all \(i\in\calS\), then the resulting approximation to the generator \(\mathbb B\) is a square matrix of dimension \(pKN + N\).\footnote{There are \(pK\) basis functions for each phase which approximate the solution on the interior of the domain, hence \(pKN\) basis functions in total, plus an additional \(N\) dimensions to capture boundary conditions.} In this sense each approximation scheme leads to matrices of the same order (although not necessarily the same number of non-zero elements).

We investigate two slope-limited schemes. In the scheme we refer to the \emph{DG-lim} scheme, we use a DG scheme of dimension \(p\) and apply the Generalised MUSCL slope limiter as described in Section~\ref{sec: slope limiting}. In the \emph{DG-lin-lim} scheme, we follow the construction described in Section~\ref{sec: anoth pos pres}, which is effectively a DG scheme with cells of width \(2\Delta/(p+1)\), for \(p=1,3,5,...,21\), and a basis of dimension \(2\) (linear) on each cell, with an upwind flux, and we apply the Generalised MUSCL limiter. Thus, the DG-lin-lim scheme uses \(p+1\) basis functions to represent the solution on each of the original cells \(\calD_k\), and we define the dimension of the scheme to be \(p+1\). In the presence of discontinuities, slope-limiting forces the approximation to be linear, i.e.~an order~1 polynomial, around the discontinuity. For DG schemes of dimension greater than 2 with a slope limiter and in the presence of discontinuities, the ability of the scheme to represent higher-order polynomials is not well-utilised as the slope limiter forces the solution to be linear anyway. The DG-lin-lim scheme recognises this and instead of using higher-order polynomials, uses piecewise-linear functions and smaller cell widths to approximate the solution.

The concept of slope-limiting originates in the context of numerically integrating (typically with respect to a \emph{time} variable) a so-called \emph{semi-discrete system}, as we do in Section~\ref{sec: transient approx}. However, approximating the operator \(\mathbb\Psi\) of a fluid-fluid queue amounts to solving an implicit matrix-equation and there is no concept of numerical integration with respect to time. Hence, the concept of slope-limiting, it is not directly transferable. One could process the initial condition and post-process the solution with a limiter. However, there are possibly better choices for this post-processing, such as filtering, \citep[Section~5.6.1]{nodalDGBook}. One could also view the composition of the DG operator and the limiting operator as one, but the result is a non-linear operator which is of no use for simplifying the Riccati Equation which characterises \(\mathbb \Psi\). Considering there is no theory on the use of the DG scheme to approximate \(\mathbb \Psi\) and hence the resultant approximation is not rigorously-defined, the author does not know if the use of a limiter in this context is possible.

@@

For schemes which require us to integrate over time we do not investigate the stability of the schemes with respect to the \(t\)-step-size, or the time-integration scheme itself (where required). Instead, we fix the time-integration step size for each numerical experiment at a suitably small value to obey a certain stability criterion (a CFL-like condition, \citep[Section~4.8]{nodalDGBook}). Moreover, we always implement the strong stability preserving Runge-Kutta method of order 4 with 5 stages (\cite{sr2002}, \cite{nodalDGBook}~Section~5.7, see also Section~\ref{sec: time integration}), which is claimed to introduce no more oscillations into the solution as we integrate over time.

@@

The structure of this chapter is as follows. In Section~\ref{sec: recon num} we compute approximations to various initial conditions for the different schemes and observe their performance which allows us to investigate the performance of the reconstruction methods without considering a specific model or any dynamics. In Section~\ref{sec: wave num} we investigate travelling wave problems with various initial conditions. For the travelling wave problem the dynamics are deterministic, which means that the solution is known, and we do not need to resort to simulation as a \emph{ground-truth}. Furthermore, the travelling wave problems allow us to investigate the ability of the schemes to approximate the flow of probability across cells without any stochastic dynamics. Next, we investigate a simple fluid queue with two phases. Section~\ref{sec:stat} investigates approximations to the limiting distribution of the queue, Section~\ref{sec: transient approx} investigates approximations to the transient distribution of the queue for two initial conditions, and Section~\ref{sec: hit approx} investigates approximations the first hitting time of the fluid level on the set \(\{0,1\}\). Lastly, we apply the approximation methods to a fluid-fluid queue in Section~\ref{sec: ffq num}, compute approximations to \(\mathbb \Psi\) and ultimately, approximate the distribution of \(\{(X(\zeta_{Y}(\{0\})),\varphi(\zeta_{Y}(\{0\})))\}\) at time \(\zeta_{Y}(\{0\})\) at which \(\{Y(t)\}\) first returns to \(0\).

@@

\subsection{QBD-RAP closing operators}\label{sec: closing vecs numerics}
We investigate the three closing operators introduced in \ref{sec: closing}. For convenience, we restate the three closing operators from Section~\ref{sec: closing} here. The unnormalised closing vector, 
\begin{align}
	\bs a {\bs v}(x) &= \bs a e^{\bs Sx}\bs s.\tag{\ref{eqn: alal}}
	%\\\bs a {\bs v}_-(x-y_\ell) &= \bs a e^{\bs S(x-y_{\ell})}\bs s,\label{eqn:ndaenf}
\end{align} 
A normalised version of the closing operator in Equation~(\ref{eqn:546}), with the normalised version given by 
\begin{align}
	\bs a {\bs v}(x) &= \cfrac{\bs a \left(e^{ \bs{S}x} \bs s + e^{ \bs{S}(2\Delta-x)} \bs s\right)}{1-\bs a e^{\bs S 2\Delta}\bs e}. \tag{\ref{eqn: nonlin}}
\end{align} 
We refer to the closing operator in (\ref{eqn: nonlin}) as the \emph{non-linear normalised} closing operator.\footnote{Technically we have not yet proved that the \emph{non-linear} normalised closing operator leads to a convergent approximation scheme.} The third closing vector we investigate is given by 
\begin{align}
	\bs a {\bs v}(x) &= \bs a \left(e^{ \bs{S}x} \bs s + e^{ \bs{S}(2\Delta-x)}\right)\left[I-e^{S2\Delta}\right]^{-1} \bs s,\tag{\ref{eqn:5462}}
\end{align}
which we refer to as the \emph{normalised} closing operator.

@@

	Figure~\ref{fig: fun 2 ks error qbdrap closing vecs} also shows that the non-linear normalised and normalised closing operators perform very similarly -- this is the case throughout much of this section. The difference between the non-linear normalised and normalised closing operators is how they treat mass in the tail of a matrix exponential distribution (from \(2\Delta\) onward). Intuitively, there should be very little mass in the tail of the distribution as we are using concentrated matrix exponential distributions in the reconstruction.

@@

Some insight is gained by looking at Figure~\ref{fig: pdf reconstructed} where we plot the reconstructed PDFs for the unnormalised (\ref{eqn: alal}) and normalised (\ref{eqn:5462}) closing operators for dimension 1, 3, 5 and 7, as well as the true PDF. Observing Figure~\ref{fig: pdf reconstructed} notice that the unnormalised reconstruction fails to capture the density at the left of each of the plots. This feature is due to a significant amount of mass being lost due to the truncation. In comparison, the reconstruction with the normalised closing operator is much better in this region due to the `folding' around \(\Delta\) in the construction of the closing operator. The `folding' in the normalised closing operator manifests itself as more mass at the left of these plots compared to the unnormalised reconstructions. % Recall that, as discussed in Section~\ref{}, for the reconstruction we must to allocate the initial condition to a phase with positive rate, or negative rate, so that we can use the appropriate approximation and reconstruction method as discussed in Section~\ref{sec: closing}. Here we suppose that the phase is positive, so the `folding' around \(\Delta\) in the normalised closing operators appears appears at the left of the plots. In general, reconstructions via the unnormalised closing operator typically underestimate the value of the function in this region.

\begin{figure}[h]
	\centering
	\includegraphics[width=\textwidth,trim={0cm 1.25cm 0cm 1.25cm},clip]{chapter6/figs/qbdrap_closing_vec/fun4/pdfs_formatted.pdf}
	\caption{Approximations to the PDF \(f(x)=1,\, x\in[0,1)\) (orange dotted line) for Example~\ref{ex: al}, constructed using matrix exponential distributions of various dimensions, and using the unnormalised closing operator (blue solid line) and normalised closing operator (green dash-dotted line).}
	\label{fig: pdf reconstructed}
\end{figure} 
%\exampleFloatBarrier
Figure~\ref{fig: pdf reconstructed} also shows that both closing operators fail to approximate the initial condition well at the right-hand side of the interval. A potential area for further research is to determine if there is a different reconstruction method or another closing operator which could alleviate this issue. 
\end{example}

\begin{example}\label{ex: aasmm}
	\begin{figure}[h]
		\centering
		\includegraphics[width=0.5\textwidth,trim={0.5cm 0.8cm 0.2cm 1.25cm},clip]{chapter6/figs/qbdrap_closing_vec/fun6/ks_error_formatted.pdf}%
		\includegraphics[width=0.5\textwidth,trim={0.5cm 0.8cm 0.2cm 1.25cm},clip]{chapter6/figs/qbdrap_closing_vec/fun6/l2_pdf_error_formatted.pdf}
		\caption{KS error (left) and \(L^2\) error for the PDFs (right) of Example~\ref{ex: aasmm}, for the three closing vectors considered; unnormalised (blue solid line), non-linear normalised (orange dashed line) and normalised (green dash-dotted line). The error curves for the non-linear normalised (orange) and normalised closing vectors are almost coincident. The black dotted lines are linear least-squares fits to the last 8 data points and the slopes of the least square lines are written next to the last point.}
		\label{fig: fun 6 ks error qbdrap closing vecs}
	\end{figure}
	Next we consider approximating the initial distribution with density \(-6x^2+6x\) on \([0,1)\). Observing the left-hand panel of Figure~\ref{fig: fun 6 ks error qbdrap closing vecs} which plots the KS error against the dimension of the matrix exponential approximation for the three closing operators, we once again see that the unnormalised closing operator performs the worst, while the performance of the two normalised closing operators is indistinguishable. However, if we instead look at the right-hand panel of Figure~\ref{fig: fun 6 ks error qbdrap closing vecs} which plots the \(L^2\) error between the reconstructed PDF and the true PDF, then the unnormalised closing operator performs better than the two normalised ones for dimension 5 and above.

	\begin{figure}[h]
		\centering
		\includegraphics[width=\textwidth,trim={0cm 1.25cm 0cm 1.25cm},clip]{chapter6/figs/qbdrap_closing_vec/fun6/pdfs_formatted.pdf}
		\caption{Approximations to the PDF \(f(x)=-6x^2+6x\) (orange dotted line) for Example~\ref{ex: aasmm}, constructed using matrix exponential distributions of various dimensions, and using the unnormalised closing operator (blue solid line) and normalised closing operator (green dash-dotted line).} 
		\label{fig: pdf reconstructed quadratic}
	\end{figure} 
	%\exampleFloatBarrier
The fact that the unnormalised closing operator outperforms the two normalised ones can be explained by the `folding' of the normalised operators around \(\Delta\). In Figure~\ref{fig: pdf reconstructed quadratic} we plot the reconstructions obtained using the unnormalised and normalised closing operators, along with the true PDF, \(-6x^2+6x\). In Figure~\ref{fig: pdf reconstructed quadratic} at the left of the plots, we observe that the normalised closing operator over-estimate the density function, whereas the unnormalised closing operator looks to be doing better. The `folding' in the normalised closing operator manifests itself as more mass at the left of these plots compared to the unnormalised closing operator. 
\end{example}

\begin{example}\label{ex: kjfk}
	\begin{figure}[h]
		\centering
		\includegraphics[width=0.5\textwidth,trim={0.5cm 0.8cm 0.2cm 1.25cm},clip]{chapter6/figs/qbdrap_closing_vec/fun7/ks_error_formatted.pdf}%
		\includegraphics[width=0.5\textwidth,trim={0.5cm 0.8cm 0.2cm 1.25cm},clip]{chapter6/figs/qbdrap_closing_vec/fun7/l2_pdf_error_formatted.pdf}
		\caption{KS error (left) and \(L^2\) error of the PDF (right) for Example~\ref{ex: kjfk} for the three closing vectors considered; unnormalised (blue solid line), non-linear normalised (orange dashed line) and normalised (green dash-dotted line). The error curves for the non-linear normalised (orange) and normalised closing vectors are almost coincident. The black dotted lines are linear least-squares fits to the last 8 data points and the slopes of the least square lines are written next to the last point.}
		\label{fig: fun 7 ks error qbdrap closing vecs}
	\end{figure}
	%\exampleFloatBarrier
	Lastly, we consider approximating the initial distribution with PDF \(3e^{-3x}/(1-e^{-3})\). This density function is at a maximum at the left of the region. Considering what we have learnt so far about the unnormalised operator underestimating in this region, we expect that the unnormalised closing operator will perform relatively poorly in this case. Indeed, observing Figures~\ref{fig: fun 7 ks error qbdrap closing vecs} we see that this is the case; the normalised closing operators perform relatively well compared to the unnormalised closing operator, as measured by both error metrics (KS statistic and \(L^2\) norm on the PDF). Plots of the PDFs (omitted), confirm that this is due to the loss of mass at the left-hand side of the region due to the truncation. 
\end{example}

Table~\ref{table: error rates for QBD-RAP closing} summarises the rates of convergence of the \(L^2\) error on the PDFs for unnormalised and normalised closing operators in the examples above (i.e.~the slopes of the black dotted lines in the error plots). In general, the errors of the approximations obtained using normalised closing operators converge at a rate approximately as fast, or faster, than the approximations obtained using the unnormalised closing operator.
\begin{table}
	\centering
	\begin{tabular}{c|c|c|c}
			Example & PDF on \([0,1)\) & \(\beta_1\), unnormalised & \(\beta_1\), normalised \\\hline
			\ref{ex: another example} & \(2\times 1(x<0.5)\) & -1.07 & -0.98 \\
			\ref{ex: al} & 1 & -0.99 & -0.91 \\
			\ref{ex: aasmm} & \(-6x^2+6x\) & -2.72 & -2.57 \\ 
			\ref{ex: kjfk} & \(3e^{-3x}/(1-e^{-3})\) & -1.11 & -2.35 
	\end{tabular}
	\caption{\label{table: error rates for QBD-RAP closing} Estimated rates of convergence, \(\beta_1\), of the \(L^2\) error on the PDFs for unnormalised and normalised closing operators in the examples above (i.e.~the slopes of the black dotted lines in the error plots).}
\end{table}

@@

\begin{example}\label{ex: exa}
	\begin{figure}[h]
		\centering
		\includegraphics[width=0.5\textwidth,trim={0.5cm 0.8cm 0.2cm 1.25cm},clip]{chapter6/figs/comp/fun1/meshs_ks_error_formatted.pdf}%
		\includegraphics[width=0.5\textwidth,trim={0.5cm 0.8cm 0.2cm 1.25cm},clip]{chapter6/figs/comp/fun1/meshs_l1_cdf_error_formatted.pdf}
		\caption{KS error (left) and \(L^1\) error of the CDF (right) of Example~\ref{ex: exa} for the DG (blue solid line), uniformisation (orange dashed line) and QBD-RAP (purple dashed line) schemes. The black dotted lines are linear least-squares fits to the last 8 data points and the slopes of the least square lines are written next to the last point.}
		\label{fig: fun 1 comp} 
	\end{figure}
	First we consider the initial condition with CDF \(1(x\geq 0.5)\), that is, a point mass of \(1\) at \(0.5\). This distribution does not have a PDF, so we compare the CDFs only. In Figure~\ref{fig: fun 1 comp} we plot the KS metric (left) and \(L^1\) metric (right) between the true CDF and the reconstructed approximations. Observing the KS metric, it appears that none of the schemes converge and the KS error sits around 0.5 for all schemes. This reflects the fact that convergence in distribution implies point-wise convergence of the CDFs except, perhaps, at points of discontinuity. None of the schemes converge at the discontinuity at \(x=0.5\). Observing the \(L^1\) error between the true CDF and reconstructed approximation (which is the area between the two CDFs) we now see that the schemes appear to show the convergent behaviour we expect. Here, the uniformisation scheme appears to perform the best, while the QBD-RAP scheme performs the worst.
